Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-02-08 Manuscript/Manuscript.Rmd
Add complete data estimates for popularity case study

# functions
`%nin%` <- Negate(`%in%`)
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
# icc for imputed data as function in mice
miceadds::source.all(path = "../R")

@@

@@

# load data 
# popmis <- mice::popmis 
#readRDS("../Data/popNCR2.RDS") %>% 
  #filter(class %in% clusters) %>% 
  #mutate(class = as.factor(class))
popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE) %>% 
  mutate(school = SCHOOL,
         popular = as.numeric(POPULAR),
         sex = SEX,
         texp = TEXP,
         .keep = "none")
comp_est <- lme4::lmer(popular ~ 1 + sex + texp + (1 + sex | school), popcomp) %>% broom.mixed::tidy()
```

@@

The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction. The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`. The regression equation and `lme4` notation for this model are

@@

Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data. The data is created in such a way that the clustering variable `school` explains quite some variance in the outcome variable `popular`. We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`. We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effect of `sex` on `popular` $\gamma_{10} = 0.843$.

@@

We'll now use `school` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16]. But at least, it includes some multilevel aspect. This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende2016]. Colloquially, this is 'multilevel imputation for dummies'.

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(-2, 2, 2, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "2l.norm", "",  "")
imp_norm_2l <-
  mice(
    popmis %>% mutate(school = as.integer(school)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

>>>>

# functions
`%nin%` <- Negate(`%in%`)
round3 <- function(x){format(round(x, 3), nsmall = 3)}
icc <- function(formula, data){multilevel::ICC1(aov(formula, data))}
# icc for imputed data as function in mice
miceadds::source.all(path = "../R")

@@

<!-- Table 7.1: Questions to gauge the complexity of a multilevel imputation task. -->
<!-- 1.	Will the complete-data model include random slopes? -->
<!-- 2.	Will the data contain systematically missing values? -->
<!-- 3.	Will the distribution of the residuals be non-normal? -->
<!-- 4.	Will the error variance differ over clusters? -->
<!-- 5.	Will there be small clusters? -->
<!-- 6.	Will there be a small number of clusters? -->
<!-- 7.	Will the complete-data model have cross-level interactions? -->
<!-- 8.	Will the dataset be very large? (fimd, section 7.3)-->

@@

# load data 
# popmis <- mice::popmis 
#readRDS("../Data/popNCR2.RDS") %>% 
  #filter(class %in% clusters) %>% 
  #mutate(class = as.factor(class))
popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE, use.value.labels = FALSE) %>% 
  mutate(school = SCHOOL,
         popular = POPULAR,
         sex = SEX,
         texp = TEXP,
         .keep = "none")
est_comp <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popcomp) 
# %>% 
#   broom.mixed::tidy(conf.int = TRUE) 
# %>% 
  # mutate(
  #   term = factor(term, levels = c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation"), labels = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor. cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error"), ordered = TRUE))
```

@@

The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction. The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`. The regression equation^[add the 'level notation' (Bryk and Raudenbush, 1992) and/or matrix notation ('linear mixed effects model'; Laird and Ware, 1982) too?] and `lme4` notation for this model are

@@

Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data. The data is created in such a way that the clustering variable `school` explains quite some variance in the outcome variable `popular`. We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`. We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effects:

```{r echo=FALSE}
# lvl <- c("(Intercept)", "sex", "texp", "sex:texp", "sd__(Intercept)", "cor__(Intercept).sex", "sd__sex", "sd__Observation")
# lbl = c("Intercept", "Sex", "Teacher experience", "Cross-level interaction", "SD cluster-level residual error", "Cor. cluster-level error and slopes sex", "SD residual slopes sex", "SD unit-level residual error")
est_comp %>% 
  broom.mixed::tidy("fixed", conf.int = TRUE) %>% 
  mutate(term = term,
         estimate  = paste0(
           round3(estimate), " [", 
           round3(conf.low), ", ", 
           round3(conf.high), "]"), 
         .keep = "none")

est_comp %>% 
  broom.mixed::tidy("ran_pars") %>% 
  mutate(term = term, #factor(term, levels = unique(.$term), labels = c()),
         estimate  = round3(estimate), 
         .keep = "none")
```

@@

We'll now use `school` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16]. But at least, it includes some multilevel aspect. This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16]. Colloquially, this is 'multilevel imputation for dummies'.

@@

"Van Buuren (2011) considered the homoscedastic linear mixed model as invalid for imputing incomplete predictors, and investigated only the 2l.norm method, which allows for heterogeneous error variances" [@buur18].

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(-2, 2, 2, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "2l.norm", "",  "")
# meth <- c("", "2l.pmm", "",  "")
imp_norm_2l <-
  mice(
    popmis %>% mutate(school = as.integer(school)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

### 2022-02-08 Manuscript/Manuscript.tex
Add complete data estimates for popularity case study

The analysis model corresponding to this dataset is multilevel
regression with random intercepts, random slopes and a cross-level
interaction. The outcome variable is \texttt{popular}, which is
predicted from the unit-level variable \texttt{sex} and the
cluster-level variable \texttt{texp}. The regression equation and
\texttt{lme4} notation for this model are

@@

Since the data is simulated and the missingness is induced, we can
compare our inferences after imputation to the true complete data. The
data is created in such a way that the clustering variable
\texttt{school} explains quite some variance in the outcome variable
\texttt{popular}. We express this using the intraclass correlation, ICC
\(=\) 0.58. We'll evaluate the ICC after each missing data strategy, and
compare the estimated fixed effect of \texttt{sex} on \texttt{popular}
\(\gamma_{10} = 0.843\).

@@

We'll now use \texttt{school} as a predictor to impute all other
variables. This is still not recommended practice, since it only works
under certain circumstances and results may be biased
\citep{drec15, ende16}. But at least, it includes some multilevel
aspect. This method is also called `fixed cluster imputation', and uses
N-1 indicator variables representing allocation of N clusters as a fixed
factor in the model \citep{reit06, ende2016}. Colloquially, this is
`multilevel imputation for dummies'.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(-2, 2, 2, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "2l.norm", "",  "")
R> imp_norm_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

>>>>

The analysis model corresponding to this dataset is multilevel
regression with random intercepts, random slopes and a cross-level
interaction. The outcome variable is \texttt{popular}, which is
predicted from the unit-level variable \texttt{sex} and the
cluster-level variable \texttt{texp}. The regression equation\footnote{add
  the `level notation' (Bryk and Raudenbush, 1992) and/or matrix
  notation (`linear mixed effects model'; Laird and Ware, 1982) too?}
and \texttt{lme4} notation for this model are

@@

Since the data is simulated and the missingness is induced, we can
compare our inferences after imputation to the true complete data. The
data is created in such a way that the clustering variable
\texttt{school} explains quite some variance in the outcome variable
\texttt{popular}. We express this using the intraclass correlation, ICC
\(=\) 0.58. We'll evaluate the ICC after each missing data strategy, and
compare the estimated fixed effects:

\begin{CodeChunk}
\begin{CodeOutput}
# A tibble: 4 x 2
  term        estimate                 
  <chr>       <chr>                    
1 (Intercept) " 3.314 [ 2.998,  3.629]"
2 sex         " 1.330 [ 1.069,  1.590]"
3 texp        " 0.110 [ 0.090,  0.130]"
4 sex:texp    "-0.034 [-0.051, -0.017]"
\end{CodeOutput}
\begin{CodeOutput}
# A tibble: 4 x 2
  term                 estimate
  <chr>                <chr>   
1 sd__(Intercept)      0.642   
2 cor__(Intercept).sex 0.077   
3 sd__sex              0.476   
4 sd__Observation      0.626   
\end{CodeOutput}
\end{CodeChunk}

@@

We'll now use \texttt{school} as a predictor to impute all other
variables. This is still not recommended practice, since it only works
under certain circumstances and results may be biased
\citep{drec15, ende16}. But at least, it includes some multilevel
aspect. This method is also called `fixed cluster imputation', and uses
N-1 indicator variables representing allocation of N clusters as a fixed
factor in the model \citep{reit06, ende16}. Colloquially, this is
`multilevel imputation for dummies'.

@@

``Van Buuren (2011) considered the homoscedastic linear mixed model as
invalid for imputing incomplete predictors, and investigated only the
2l.norm method, which allows for heterogeneous error variances''
\citep{buur18}.

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(-2, 2, 2, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "2l.norm", "",  "")
R> # meth <- c("", "2l.pmm", "",  "")
R> imp_norm_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}
